Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-02-10 Manuscript/HI_Model_ms.Rmd
updated manuscript files

### *Carbon stocks and fluxes*
Terrestrial ecosystems of the seven main Hawaiian Islands stored an estimated 316 Tg of carbon at the beginning of the simulation period in 2010, with 58% in soil organic matter, 22% in living biomass, and 20% in surface dead organic matter (litter and dead wood; figure 2a). Ecosystems accumulated carbon in all scenarios but at different rates, with trajectories shaped primarily by climate change and to a lesser extent by land-use change. The highest and most consistent projected accumulation of ecosystem carbon occured under the combination of low radiative forcing and low land use change, yielding a ~15% increase in ecosystem carbon to an average of 363 Tg by 2100 (figure 2a). In contrast, high radiative forcing and high land use change resulted in the lowest ecosystem carbon gain, reaching a peak of ~332 Tg in 2063 but declining to 327 Tg in 2100, resulting in a net increase of only 3% by the end of the simulation period (figure 2a). Ecosystem carbon accumulation was driven exclusively by increasing soil organic carbon across all four scenarios, all other stocks declined over time (figure 2b).

@@

Net primary production (NPP) for the seven main Hawaiian Islands declined across all four scenarios, driven primarily by climate change and to a lesser extent by land use change (Fig. 3). The combination of high radiative forcing (RCP 8.5) and high land-use change led to the steepest decline in NPP over time, driven by intense long-term drying on the leeward sides of islands under RCP 8.5 (figure S4) and sustained losses of forest and shrubland land area in the high land-use scenario (figure S5). In contrast, climate change led to increased heterotrophic respiration (R~h~) over time, such that more intense warming under RCP 8.5 (figure S4) resulted in R~h~ being ~3% higher by 2100 than under RCP 4.5 (figure 3).  Land-use change substantially reduced R~h~ in the high land-use scenario (figure 3) because of long-term reductions in forest and shrubland land area (figure S5), similar to trends in NPP. Transition-triggered carbon fluxes to the atmosphere from land use, land-use change, and wildfire were largely independent of changes in climate, stabilizing by mid-century at an average of ~0.4 Tg y^-1^ in the high land-use scenario and ~0.2 Tg y^-1^ in the low land-use scenario (figure 3). Uncertainty around transition-triggered carbon fluxes were higher in the high land-use scenario, driven primarily by greater variability in wildland fire probabilities.

@@

We estimated that terrestrial ecosystems of the Hawaiian Islands have been a consistent net sink for atmospheric carbon over the last decade (figure 4). Net biome productivity (NBP) averaged 0.64 TgC y^-1^ and ranged from 0.46 to 0.88 TgC y^-1^ across all scenarios for the time period 2011-2019. Based on these LUCAS model estimates, Hawaiian terrestrial ecosystems offset approximately 13% of 2015 statewide CO~2~ emissions from energy production and transportation (5.04 TgC), the State of Hawaiʻi's largest source of greenhouse gas emissions [@hawaii_inventory_2019]. Future projections indicate Hawaiian terrestrial ecosystems will continue to be a net sink for atmospheric carbon if global CO~2~ emissions peak around 2040 and then decline (RCP 4.5), and that reducing the intensity of future land use change in Hawaiʻi will further enhance carbon sink strength. If, however, global emissions continue to rise throughout the 21st century (RCP 8.5), our projections indicate Hawaiian ecosystems will transition from a net sink to a net source of CO~2~ to the atmosphere, with high intensity land-use change accelerating this transition and exacerbating net carbon loss. Our model results also indicate that projections of ecosystem carbon balance are highly sensitive to the introduction of a CFE. Even a 5% increase in NPP per 100ppm increase in atmospheric CO~2~ was sufficient to maintain Hawaiian Island ecosystems as a net carbon sink under the high radiative forcing and high land-use scenario for the entire simulation period. Reconciling the high uncertainty surrounding the response of net photosynthesis to rising atmospheric CO~2~ is essential to more realistically constrain simulation model projections.

Our NBP estimates of ~ 0.64 TgC y^-1^ for the recent past (2011-2019) agree well with those from a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 TgC in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019].

>>>>

### *Carbon stocks and fluxes*
Terrestrial ecosystems of the seven main Hawaiian Islands stored an estimated 316 Tg of carbon at the beginning of the simulation period in 2010 (figure 2a), with 58% in soil organic matter, 22% in living biomass, and 20% in surface dead organic matter (litter and dead wood; figure 2b). Ecosystems accumulated carbon in all scenarios but at different rates, with trajectories shaped primarily by climate change and to a lesser extent by land-use change. The highest and most consistent projected accumulation of ecosystem carbon occured under the combination of low radiative forcing and low land use change, yielding a ~15% increase in ecosystem carbon to an average of 363 Tg by 2100 (figure 2a). In contrast, high radiative forcing and high land use change resulted in the lowest ecosystem carbon gain, reaching a peak of ~332 Tg in 2063 and a decline to 327 Tg in 2100, resulting in a net increase of only 3% by the end of the simulation period (figure 2a). Ecosystem carbon accumulation was driven exclusively by increasing soil organic carbon across all four scenarios, all other stocks declined over time (figure 2b).

@@

Net primary production (NPP) for the seven main Hawaiian Islands declined across all four scenarios, driven primarily by climate change and to a lesser extent by land use change (Fig. 3). The combination of high radiative forcing (RCP 8.5) and high land-use change led to the steepest decline in NPP over time, driven by intense long-term drying on the leeward sides of islands under RCP 8.5 (figure S4) and sustained losses of forest and shrubland land area in the high land-use scenario (figure S5). In contrast, climate change led to increased heterotrophic respiration (R~h~) over time, such that more intense warming under RCP 8.5 (figure S4) resulted in R~h~ being ~ 3% higher by 2100 than under RCP 4.5 (figure 3). Heterotrophic respiration declined substantially over time in the high land-use scenario (figure 3) because of long-term reductions in forest and shrubland land area (figure S5), similar to trends in NPP. Transition-triggered carbon fluxes to the atmosphere from land use, land-use change, and wildfire were largely independent of changes in climate, stabilizing by mid-century at an average of ~0.4 Tg y^-1^ in the high land-use scenario and ~0.2 Tg y^-1^ in the low land-use scenario (figure 3). Uncertainty around transition-triggered carbon fluxes were higher in the high land-use scenario, driven primarily by greater variability in wildland fire probabilities.

@@

We estimated that terrestrial ecosystems of the Hawaiian Islands have been a consistent net sink for atmospheric carbon over the last decade (figure 4). For the time period 2011-2019, net biome productivity (NBP) averaged 0.64 TgC y^-1^ and ranged from 0.46 to 0.88 TgC y^-1^ across all scenarios. Based on this mean annual NBP estimate, Hawaiian terrestrial ecosystems offset approximately 13% of 2015 statewide CO~2~ emissions from energy production and transportation (5.04 TgC), the State of Hawaiʻi's largest source of greenhouse gas emissions [@hawaii_inventory_2019]. Future projections indicate Hawaiian terrestrial ecosystems will continue to be a net sink for atmospheric carbon if global CO~2~ emissions peak around 2040 and then decline (RCP 4.5), and that carbon sink strength can be further enhanced by reducing the intensity and extent of future land use change. If, however, global CO~2~ emissions continue to rise throughout the 21st century (RCP 8.5), our projections indicate Hawaiian ecosystems will transition from a net sink to a net source of CO~2~ to the atmosphere, with high levels of land-use change accelerating this transition and exacerbating net carbon loss. Our model results also indicate that projections of ecosystem carbon balance are highly sensitive to the introduction of a CFE. Even a 5% increase in NPP for every 100ppm increase in atmospheric CO~2~ was sufficient to completely offset all other carbon losses induced by the high radiative forcing and high land use scenario, maintaining Hawaiian Island ecosystems as a net carbon sink for the entire simulation period instead of transitioning to a net carbon source by mid-century. Reconciling the high uncertainty surrounding the response of net photosynthesis to rising atmospheric CO~2~ is essential to more realistically constrain simulation model projections.

### *Drivers of ecosystem carbon balance*

### *Comparison to other studies*
There are very few estimates of contemporary ecosystem carbon balance for the main Hawaiian Islands, and even fewer model projections of future ecosystem carbon balance.  Our mean annual NBP estimate of 0.64 TgC y^-1^ for the period 2011-2019 agrees well with a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 TgC in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019].

### *Limitations of this study*
